A spatial analysis of the environmental correlates of COVID-19 incidence in Spain

### 2020-04-18 Environmental-Correlates-of-COVID19-Spain/Environmental-Correlates-of-COVID19-Spain.Rmd
Edits to Context and Methods

For a better interpretation of the results of our analysis is necessary to know the chronology of the epidemic outbreak in Spain. This subsection presents a brief time line of COVID-19.

The first reported case of COVID-19 in Spain was on January 31th, were a German tourist in the Canary Islands was positively tested, but the first not imported case was diagnostic on February 27th in Sevilla province (Andalusia). The epidemic was growing slowly at the beginning but in March 11th the World Health Organization (WHO) declared officially the pandemic. This declaration was a break point and on March 13th the Spanish National Government declare the "Emergency Status". At this date, Spain have been declared 4209 cases of COVID-19 (mostly of cases, 1990, in Madrid). Only essential services (e.g. food, health) and some economic subsector of industry and construction could to maintain activity. On March 17th Spain closes its lands borders and only allows entry to nationals and residents. From March 30th to April 12th (April 10th-12th Easter) lockdown of the Spanish people was hardened and only essential activities were allowed.

@@

Our dataset includes information about the daily number of cases of COVID-19 reported in Spain at the provincial level (NUTIII in Eurostat terminology) for the period between March 13th and April 11th, inclusive. This gives a temporal coverage of 30 days and spatial coverage of 50 units. For our purposes, we consider positive cases reported, but excluding symptomatic cases diagnosed by a doctor without a Polymerase Chain Reaction (PCR) test. The Spanish National Government publishes periodic updates at the regional level (NUTII) and the information is also released at the provincial level as part of a collaborative project [**by whom?**]. This information is compiled from several sources, mainly the official web pages of the Spanish regional goverments. In addition, we consider two sets of explanatory variables. The first one, and the focus of this research, is set of two environmental variables, namely temperature and humidity, which are collected from official sources (i.g., AEMET, the state meteorology agency, and MAPA, the ministry of agriculture, fisheries, and food). The second set provides some relevant controls for multivariate analysis, and refers to economic and demographic attributes of the province (also collected from official sources, i.e., INE, the national statistics institute). Table \ref{tab:descriptive-statistics} shows the descriptive statistics and the provenance of the data.

```{r descriptive-statistics, echo=FALSE}
data.frame(Variable = c("COVID-19 Incidence",
                        "GDPpc", 
                        "Older", 
                        "Population Density",
                        "Mean Temperature",
                        "Humidity"),
           Note = c("Incidence in reported cases of SARS-19 per 100,000 people",
                    "GDP per capita in €1,000s",
                    "Percentage of people aged 65 and older in the province",
                    "Population density in the province in people per sq.km",
                    "Mean temperature in province by date, in C",
                    "Relative humidity in province by date"),
           Min = c(min(covid19_spain$Incidence),
                   min(provinces_spain$GDPpc),
                   min(provinces_spain$Older),
                   min(provinces_spain$Density),
                   min(covid19_spain$Mean_Temp),
                   min(covid19_spain$Humidity)),
           Mean = c(mean(covid19_spain$Incidence),
                   mean(provinces_spain$GDPpc),
                   mean(provinces_spain$Older),
                   mean(provinces_spain$Density),
                   mean(covid19_spain$Mean_Temp),
                   mean(covid19_spain$Humidity)),
           Max = c(max(covid19_spain$Incidence),
                   max(provinces_spain$GDPpc),
                   max(provinces_spain$Older),
                   max(provinces_spain$Density),
                   max(covid19_spain$Mean_Temp),
                   max(covid19_spain$Humidity)),
           SD = c(sd(covid19_spain$Incidence),
                   sd(provinces_spain$GDPpc),
                   sd(provinces_spain$Older),
                   sd(provinces_spain$Density),
                   sd(covid19_spain$Mean_Temp),
                   sd(covid19_spain$Humidity)),
           Source = c("Montera34",
                      "INE",
                      "INE",
                      "INE",
                      "AEMET",
                      "MAPA")) %>%
  kable("latex",
        #"html",
        booktabs = TRUE,
        digits = 2,
        caption = "\\label{tab:descriptive-statistics}Descriptive statistics") %>%
  kable_styling(latex_options = c("striped", "scale_down")) %>%
  #kable_styling(bootstrap_options = c("striped", "condensed")) %>%
  column_spec(2, width = "15em") %>%
  column_spec(7, width = "5em") %>%
  footnote(general = c("Montera34: https://code.montera34.com:4443/numeroteca/covid19/-/blob/master/README_providencialdata19.md", 
                       "INE (Instituto Nacional de Estadistica): https://www.ine.es/",
                       "AEMET (Agencia Estatal de Meteorologia): http://eportal.mapa.gob.es",
                       "MAPA (Ministerio de Agricultura, Pesca y Alimentacion): http://eportal.mapa.gob.es"))
```

@@

The Seemingly Unrelated Regression equations model (SUR from now on) is a multivariate econometric formulation used in different fields when the data set is organized as cross-sections for different time periods. The basis of this approach is well-known since the initial works of Zellner [-@Zellner1962efficient] and now is a popular methodology included in several Econometrics textbook [e.g., @Greene2003econometric]. To our knowledge, @Anselin1988spatial was the first text that introduced the term spatial SUR in reference to a model made of ‘an equation for each time period, which is estimated for a cross section of spatial units’ (p. 141). From this milestone, a huge researches used this methodology in spatial framework. Rey and Montouri [-@Rey1999us], Fingleton [-@Fingleton2007multi], Lauridsen et al. [-@Lauridsen2010spatiotemporal], Le Gallo and Dall’Erba [-@Legallo2006evaluating], and Lopez et al. [-@Lopez2017spatial] are well-known examples in this line.

The classical SUR model without spatial effects (from here, SUR-SIM) is a stack of equations as follows:

@@

<!--Equation for README.md generated using this app https://alexanderrodin.com/github-latex-markdown/-->
<!--
![\begin{equation} \label{eq:sur-sim} \begin{bmatrix} y_1 \\ y_2 \\ \vdots \\ y_T \end{bmatrix} = \begin{bmatrix} X_1 & 0 & \cdots & 0 \\ 0 & X_2 & \cdots & 0 \\ \vdots & \vdots & \ddots & \vdots \\ 0 & 0 & \cdots & X_T \end{bmatrix} \ \begin{bmatrix} \beta_1 \\ \beta_1 \\ \vdots \\ \beta_T \end{bmatrix} + \begin{bmatrix} \epsilon_1 \\ \epsilon_2 \\ \vdots \\ \epsilon_T \end{bmatrix} \end{equation}](https://render.githubusercontent.com/render/math?math=%5Cbegin%7Bequation%7D%20%5Clabel%7Beq%3Asur-sim%7D%20%5Cbegin%7Bbmatrix%7D%20y_1%20%5C%5C%20y_2%20%5C%5C%20%5Cvdots%20%5C%5C%20y_T%20%5Cend%7Bbmatrix%7D%20%3D%20%5Cbegin%7Bbmatrix%7D%20X_1%20%26%200%20%26%20%5Ccdots%20%26%200%20%5C%5C%200%20%26%20X_2%20%26%20%5Ccdots%20%26%200%20%5C%5C%20%5Cvdots%20%26%20%5Cvdots%20%26%20%5Cddots%20%26%20%5Cvdots%20%5C%5C%200%20%26%200%20%26%20%5Ccdots%20%26%20X_T%20%5Cend%7Bbmatrix%7D%20%5C%20%5Cbegin%7Bbmatrix%7D%20%5Cbeta_1%20%5C%5C%20%5Cbeta_1%20%5C%5C%20%5Cvdots%20%5C%5C%20%5Cbeta_T%20%5Cend%7Bbmatrix%7D%20%2B%20%5Cbegin%7Bbmatrix%7D%20%5Cepsilon_1%20%5C%5C%20%5Cepsilon_2%20%5C%5C%20%5Cvdots%20%5C%5C%20%5Cepsilon_T%20%5Cend%7Bbmatrix%7D%20%5Cend%7Bequation%7D)
-->

\noindent where $y_{t}=(y_{1t},...,y_{Nt})$ is a $N \times 1$ vector, and in our case $y_{st}$ is the incidence ratio in the province $s$ ($s=1,...,N$) the day $t$ $(t=1,...,T)$; $X_t=(X^1,...,X^{k_t})$ is a $N \times k_t$ matrix of the $k_t$ independent variables, $X_i=(X^i_{st})$; $\beta_t=(\beta_{1t},...,\beta_{Nt})$ is a vector of coefficients and $\epsilon_t=(\epsilon_{1t},...,\epsilon_{Nt})$ is the vector of residuals.

The more relevant characteristic of a SUR model is the dependence structure among the vectors of residuals, namely:

@@

<!--
![\begin{equation} \label{eq:sur-err} E\[\epsilon_t \epsilon'_{t'}\]=\sigma_{tt'} \end{equation} ](https://render.githubusercontent.com/render/math?math=%5Cbegin%7Bequation%7D%20%5Clabel%7Beq%3Asur-err%7D%20E%5B%5Cepsilon_t%20%5Cepsilon'_%7Bt'%7D%5D%3D%5Csigma_%7Btt'%7D%20%5Cend%7Bequation%7D%20)
-->

Note that this specification is very flexible, in that it allows changes in the coefficients $\beta_{it}$ in order to modulate the effect of $X^i_{.t}$ on $y_t$. This flexibility can be reduced and it is posible to impose restrinctions considering some $\beta$ coefficients as being constant over time. In this case, we can reformulate the coefficients expression of $\beta_t = (\beta_{1}, \cdots, \beta_{r-1}, \beta_{r}, \beta_{r+1}, \cdots, \beta_{Nt})$ considering the first $r$ coefficients contanst.

Equation \ref{eq:sur-sim}) can be rewriten in compact form:

@@

<!--
![\begin{equation} y = X \beta + \epsilon \end{equation}](https://render.githubusercontent.com/render/math?math=%5Cbegin%7Bequation%7D%20y%20%3D%20X%20%5Cbeta%20%2B%20%5Cepsilon%20%5Cend%7Bequation%7D)
-->

@@

As is the case of cross-sectional data, it is possible identify spatial autocorrelation in the residuals of Model \ref{eq:sur-sim}, and several Lagrange Multipliers tests have been developed to test the null hypothesis of spatial independence [see @Lopez2014spatial]. When the null hypothesis is rejected, there are alternative several specifications that have been propose to include spatial effects [@Anselin1988spatial; see also @Anselin2016estimation]. In this paper we consider a spatial SUR model that incorporates a spatial lag of the dependent variable as an explanatory factor. The stack expresion for the spatial lag SUR model (SUR-SLM),

@@

<!--
![\begin{equation} \label{eq:sur-slm} \bf{A}y = X \beta + \epsilon \\ \epsilon =N(0,\Omega) \end{equation}](https://render.githubusercontent.com/render/math?math=%5Cbegin%7Bequation%7D%20%5Clabel%7Beq%3Asur-slm%7D%20%5Cbf%7BA%7Dy%20%3D%20X%20%5Cbeta%20%2B%20%5Cepsilon%20%5C%5C%20%5Cepsilon%20%3DN(0%2C%5COmega)%20%5Cend%7Bequation%7D)
-->

\noindent where A =$I_{TN}-\bf{\Lambda} \otimes W$ with $\bf{\Lambda} = diag(\lambda_1, \cdots, \lambda_T)$.

This specification assumes that incidence in a province ($y_st$) at time $t$ is partially determined by the weighted average ($Wy_{st}$) of incidence in neighbouring provinces. Parameters of spatial dependence are estimated for each time period $\lambda_t$ and identifies the intensity and the sign of the impacts of neighbourhood. It is possible test the null hypotheis of identtical levels of spatial dependence ($\lambda_i=\lambda_j, \forall i,j$). The correspond Wald test is available in the R package **spsur**

@@

Figure \ref{fig:weekly-average-incidence-map} shows the geographical variation in the incidence of the disease, as well as its temporal progression in weekly averages. Our analysis begins on March 13, which is the first date when every province had reported at least one case of COVID-19. It can be seen that the highest incidence at this early date was in the provice of Álava, in the North of Spain. While not the most populous province, with a population of only 331,549, Álava has the highest GDP per capita of all provinces. Vitoria, its main city, is the capital of the Basque country and has been the focus of efforts to develop a "Global Basque City" [@Meijers2008strategic], along with San Sebastian and Bilbao. The other early focus of the pandemic in Spain was Madrid, which is the most populous province in the country and has the second highest GDP per capita after Álava. The early outbreaks in these two provinces can be traced throughout the progression of the pandemic over time, although by the end of the period under consideration, other provinces had matched and even surpased their incidence rates, including Segovia and Soria to the north of Madrid, and Ciudad Real and Albacete to the south.

@@

```{r model1-8-day-moving-average, include=FALSE}
# Model with a lagged 8-day moving average of climatic variables:
sur.slm_lag8 <- spsur::spsurtime(formula = formula_lag8, 
                                 data=GPanel, 
                                 time = GPanel$Date, 
                                 type = "slm", 
                                 fit_method = "3sls", 
                                 listw=  listw,
                                 R = R2,
                                 b = b2)
#summary(sur.slm_lag8)
#print(paste("Pooled R^2 = ", sur.slm_lag8$R2[1]))
```

```{r model2-11-day-moving-average, include=FALSE}
# Model with 11-day moving average of climatic variables:
sur.slm_lag11 <- spsur::spsurtime(formula = formula_lag11, 
                                  data=GPanel, 
                                  time = GPanel$Date, 
                                  type = "slm", 
                                  fit_method = "3sls", 
                                  listw=  listw,
                                 R = R2,
                                 b = b2)
#summary(sur.slm_lag11)
#print(paste("Pooled R^2 = ", sur.slm_lag11$R2[1]))
```

```{r model3-11-day-weighted-moving-average, include=FALSE}
# Model with 11-day weighted moving average of climatic variables:
sur.slm_lag11w <- spsur::spsurtime(formula = formula_lag11w, 
                                   data=GPanel, 
                                   time = GPanel$Date, 
                                   type = "slm", 
                                   fit_method = "3sls", 
                                   listw=  listw,
                                 R = R2,
                                 b = b2)
#summary(sur.slm_lag11w)
#print(paste("Pooled R^2 = ", sur.slm_lag11w$R2[1]))
```

>>>>

The first reported case of COVID-19 in Spain was on January 31th, when a German tourist in the Canary Islands tested positive for the virus. However, it was still a few weeks before the first domestic case was reported, on February 27th in Sevilla province (Andalusia). In a short period of time, after this relatively slow start, the outbreak flared. By March 11th the World Health Organization (WHO) declared COVID-19 officially a pandemic. This declaration marked a turning point for the public in Spain too. As of March 13th, the number of cases of COVID-19 reported in Spain was 4,473, with a majority of cases (1,990) concentrated in Madrid: these numbers were at the time the worst outbreak in Europe, after Italy. In response to the situation, on March 13th the Spanish National Government declared a state of emergency, to go into effect on Saturday March 14th. As part of the state of emergency restrictions to most activities were imposed, with the exception of essential services (e.g. food, health) and some economic subsectors of industry and construction. A few days later, on March 17th, Spain closed its lands borders to allow entry only to returnee nationals and permanent residents. The lockdown was further hardened between March 30th and April 12th (including the Easter weekend of April 10th-12th) and during this period only essential activities were allowed. During this period, there was a dramatic reduction in overall mobility, both within provinces as between \footnote{https://www.mitma.gob.es/ministerio/covid-19/evolucion-movilidad-big-data/movilidad-provincial}.

@@

Our dataset includes information about the daily number of cases of COVID-19 reported in Spain at the provincial level (NUTIII in Eurostat terminology) for the period between March 13th and April 11th, inclusive. For our purposes, we consider positive cases reported, but excluding symptomatic cases diagnosed by a doctor without a Polymerase Chain Reaction (PCR) test. The Spanish National Government publishes periodic updates at the regional level (NUTII) and the information is also released at the provincial level as part of a collaborative project [**by whom?**]. This information is compiled from several sources, mainly the official web pages of the Spanish regional goverments. In addition, we consider two sets of explanatory variables. The first one, and the focus of this research, is set of two environmental variables, namely temperature and humidity, which are collected from official sources (i.g., AEMET, the state meteorology agency, and MAPA, the ministry of agriculture, fisheries, and food). The second set provides some relevant controls for multivariate analysis, and refers to economic and demographic attributes of the province (also collected from official sources, i.e., INE, the national statistics institute). Table \ref{tab:descriptive-statistics} shows the descriptive statistics and the provenance of the data.

The spatial and temporal coverage of the data is as follows. Our dataset begins on March 13, which is the first date when every province had reported at least one case of COVID-19, and continues until April 11, for a period of 30 days. The unit of analysis is the province. Provinces are the equivalent of states, and are embedded in Autonomous Communities. As an example, Cataluña is an Autonomous Community and consists of four provinces, namely Barcelona, Gerona, Lerida, and Tarragona. The size of the provinces is relatively large, as seen in Table \ref{tab:descriptive-statistics}. The smallest province is $1,978.12km^2$ (this is smaller than Rhode Island in the US) and the largest province is $21,767.93km^2$ (slightly smaller than New Jersey in the US). While this is a relatively large degree of spatial aggregation, reporting on COVID-19 is inconsistent at smaller geographies, or cases are not reported at that level at all. The analysis must therefore be considered ecological.

```{r descriptive-statistics, echo=FALSE}
data.frame(Variable = c("COVID-19 Incidence",
                        "Area",
                        "GDPpc", 
                        "Older", 
                        "Population Density",
                        "Mean Temperature",
                        "Humidity"),
           Note = c("Incidence in reported cases of SARS-19 per 100,000 people",
                    "Area of province in sq.km",
                    "GDP per capita in €1,000s",
                    "Percentage of people aged 65 and older in the province",
                    "Population density in the province in people per sq.km",
                    "Mean temperature in province by date, in C",
                    "Relative humidity in province by date"),
           Min = c(min(covid19_spain$Incidence),
                   min(provinces_spain %>% 
                         st_area() %>% 
                         set_units(km^2)), # Convert area from sq.m to sq.km
                   min(provinces_spain$GDPpc),
                   min(provinces_spain$Older),
                   min(provinces_spain$Density),
                   min(covid19_spain$Mean_Temp),
                   min(covid19_spain$Humidity)),
           Mean = c(mean(covid19_spain$Incidence),
                   mean(provinces_spain %>% 
                         st_area()/(1000 * 1000)), # Convert area from sq.m to sq.km
                   mean(provinces_spain$GDPpc),
                   mean(provinces_spain$Older),
                   mean(provinces_spain$Density),
                   mean(covid19_spain$Mean_Temp),
                   mean(covid19_spain$Humidity)),
           Max = c(max(covid19_spain$Incidence),
                   max(provinces_spain %>% 
                         st_area()/(1000 * 1000)), # Convert area from sq.m to sq.km
                   max(provinces_spain$GDPpc),
                   max(provinces_spain$Older),
                   max(provinces_spain$Density),
                   max(covid19_spain$Mean_Temp),
                   max(covid19_spain$Humidity)),
           SD = c(sd(covid19_spain$Incidence),
                   sd(provinces_spain %>% 
                         st_area()/(1000 * 1000)), # Convert area from sq.m to sq.km
                   sd(provinces_spain$GDPpc),
                   sd(provinces_spain$Older),
                   sd(provinces_spain$Density),
                   sd(covid19_spain$Mean_Temp),
                   sd(covid19_spain$Humidity)),
           Source = c("Montera34",
                      "INE",
                      "INE",
                      "INE",
                      "INE",
                      "AEMET",
                      "MAPA")) %>%
  kable("latex",
        #"html",
        booktabs = TRUE,
        digits = 2,
        caption = "\\label{tab:descriptive-statistics}Descriptive statistics") %>%
  kable_styling(latex_options = c("striped", "scale_down")) %>%
  #kable_styling(bootstrap_options = c("striped", "condensed")) %>%
  column_spec(2, width = "15em") %>%
  column_spec(7, width = "5em") %>%
  footnote(general = c("Montera34: https://code.montera34.com:4443/numeroteca/covid19/-/blob/master/README_providencialdata19.md", 
                       "INE (Instituto Nacional de Estadistica): https://www.ine.es/",
                       "AEMET (Agencia Estatal de Meteorologia): http://eportal.mapa.gob.es",
                       "MAPA (Ministerio de Agricultura, Pesca y Alimentacion): http://eportal.mapa.gob.es"))
```

An important aspect of working with environmental data such as temperature and humidity is the incubation period of the disease. Lauer et al. [-@Lauer2020incubation] report for the case of COVID-19 a median incubation period of 5.7 days (with a confidence interval between 4.9 to 7.8 days). The vast majority of cases (95%) develop symptoms between 2.6 days (CI, 2.1 to 3.7 days) and 12.5 days (CI, 8.2 to 17.7 days). For this reason, we judge it best to use lagged values of the environmental variables. We test different time lags as follows. We consider a lagged 8-day average, from date-minus-12 to date-minus-5 days (hereafter *lag8*). Secondly, we consider a lagged 11-day average, from date-minus-12 to date-minus-2 days (hereafter *lag11*). Finally, to account for the likely duration of incubation, we consider a lagged 11-day *weighted* average, from date-minus-12 to date-minus-2 days (hereafter *lag11w*). The weights for this variable are calculated using the parameters of the log-normal distribution reported by Lauer et al. [-@Lauer2020incubation], i.e., a log-mean of 1.621 and a log-standard deviation of 0.418. With these weights, the environmental variables at date-minus-2 and date-minus-12 days are weighted as 0.041 and 0.009 respectively, whereas the environmental variables at date-minus-5 days are weighted as 0.194.

@@

The Seemingly Unrelated Regression equations model (SUR hereafter) is a multivariate econometric model used in different fields when the structure of the data consisits of cross-sections for different time periods. The basis of this approach is well-known since the initial works of Zellner [-@Zellner1962efficient], and has become a popular methodology included in several econometrics textbook [e.g., @Greene2003econometric]. To our knowledge, Anselin [-@Anselin1988spatial] was the first author to discuss SUR from a spatial perspective. In his landmark text, Anselin discussed a model made of "an equation for each time period, which is estimated for a cross section of spatial units" (p. 141). From this milestone, a large body of research has developed to extend the classical SUR into a spatial framework [e.g., @Rey1999us; @Lauridsen2010spatiotemporal; @Legallo2006evaluating; @Lopez2017spatial].

The classical SUR model without spatial effects (hereafter, SUR-SIM) is a stack of equations as follows:

@@

\noindent where $y_{t}=(y_{1t},...,y_{Nt})$ is an $N \times 1$ vector, and in our case $y_{st}$ is the incidence ratio in province $s\text{ }s=1,...,N$) on day $t\text{ }(t=1,...,T)$; $X_t=(X^1,...,X^{k_t})$ is an $N \times k_t$ matrix of the $k_t$ independent variables, $X_i=(X^i_{st})$; $\beta_t=(\beta_{1t},...,\beta_{Nt})$ is a vector of coefficients and $\epsilon_t=(\epsilon_{1t},...,\epsilon_{Nt})$ is the usual vector of random residuals.

<!--Equations for README.md generated using this app https://alexanderrodin.com/github-latex-markdown/-->
<!--
![\begin{equation} \label{eq:sur-sim} \begin{bmatrix} y_1 \\ y_2 \\ \vdots \\ y_T \end{bmatrix} = \begin{bmatrix} X_1 & 0 & \cdots & 0 \\ 0 & X_2 & \cdots & 0 \\ \vdots & \vdots & \ddots & \vdots \\ 0 & 0 & \cdots & X_T \end{bmatrix} \ \begin{bmatrix} \beta_1 \\ \beta_1 \\ \vdots \\ \beta_T \end{bmatrix} + \begin{bmatrix} \epsilon_1 \\ \epsilon_2 \\ \vdots \\ \epsilon_T \end{bmatrix} \end{equation}](https://render.githubusercontent.com/render/math?math=%5Cbegin%7Bequation%7D%20%5Clabel%7Beq%3Asur-sim%7D%20%5Cbegin%7Bbmatrix%7D%20y_1%20%5C%5C%20y_2%20%5C%5C%20%5Cvdots%20%5C%5C%20y_T%20%5Cend%7Bbmatrix%7D%20%3D%20%5Cbegin%7Bbmatrix%7D%20X_1%20%26%200%20%26%20%5Ccdots%20%26%200%20%5C%5C%200%20%26%20X_2%20%26%20%5Ccdots%20%26%200%20%5C%5C%20%5Cvdots%20%26%20%5Cvdots%20%26%20%5Cddots%20%26%20%5Cvdots%20%5C%5C%200%20%26%200%20%26%20%5Ccdots%20%26%20X_T%20%5Cend%7Bbmatrix%7D%20%5C%20%5Cbegin%7Bbmatrix%7D%20%5Cbeta_1%20%5C%5C%20%5Cbeta_1%20%5C%5C%20%5Cvdots%20%5C%5C%20%5Cbeta_T%20%5Cend%7Bbmatrix%7D%20%2B%20%5Cbegin%7Bbmatrix%7D%20%5Cepsilon_1%20%5C%5C%20%5Cepsilon_2%20%5C%5C%20%5Cvdots%20%5C%5C%20%5Cepsilon_T%20%5Cend%7Bbmatrix%7D%20%5Cend%7Bequation%7D)

\noindent where ![$y_{t}=(y_{1t},...,y_{Nt})$](https://render.githubusercontent.com/render/math?math=%24y_%7Bt%7D%3D(y_%7B1t%7D%2C...%2Cy_%7BNt%7D)%24) is an ![$N \times 1$](https://render.githubusercontent.com/render/math?math=%24N%20%5Ctimes%201%24) vector, and in our case ![$y_{st}$](https://render.githubusercontent.com/render/math?math=%24y_%7Bst%7D%24) is the incidence ratio in province ![$s\text{ }(s=1,...,N)$](https://render.githubusercontent.com/render/math?math=%24s%5Ctext%7B%20%7D(s%3D1%2C...%2CN)%24) on day ![$t\text{ }(t=1,...,T)$](https://render.githubusercontent.com/render/math?math=%24t%5Ctext%7B%20%7D(t%3D1%2C...%2CT)%24); ![$X_t=(X^1,...,X^{k_t})$](https://render.githubusercontent.com/render/math?math=%24X_t%3D(X%5E1%2C...%2CX%5E%7Bk_t%7D)%24) is a ![$N \times k_t$](https://render.githubusercontent.com/render/math?math=%24N%20%5Ctimes%20k_t%24) matrix of the ![$k_t$](https://render.githubusercontent.com/render/math?math=%24k_t%24) independent variables, ![$X_i=(X^i_{st})$](https://render.githubusercontent.com/render/math?math=%24X_i%3D(X%5Ei_%7Bst%7D)%24); ![$\beta_t=(\beta_{1t},...,\beta_{Nt})$](https://render.githubusercontent.com/render/math?math=%24%5Cbeta_t%3D(%5Cbeta_%7B1t%7D%2C...%2C%5Cbeta_%7BNt%7D)%24) is a vector of coefficients and ![$\epsilon_t=(\epsilon_{1t},...,\epsilon_{Nt})$](https://render.githubusercontent.com/render/math?math=%24%5Cepsilon_t%3D(%5Cepsilon_%7B1t%7D%2C...%2C%5Cepsilon_%7BNt%7D)%24) is the usual vector of random residuals.
-->

A key feature of the SUR framework is the dependence structure among the vectors of random residuals, which takes the following form:

@@

Note that this specification is very flexible, since it allows the coefficients $\beta_{it}$ to vary in order to modulate the effect of $X^i_{.t}$ on $y_t$. A less flexible if more parsimonious model can be obtained by imposing restrictions to some $\beta$ coefficients to fix them across equations. This is equivalent to specifying an effect that is constant in time. In this case, we can reformulate the coefficients expression of $\beta_t = (\beta_{1}, \cdots, \beta_{r-1}, \beta_{r}, \beta_{r+1}, \cdots, \beta_{Nt})$ to restrict the first $r$ coefficients to be constant.

<!--
![\begin{equation} \label{eq:sur-err} E\[\epsilon_t \epsilon'_{t'}\]=\sigma_{tt'} \end{equation} ](https://render.githubusercontent.com/render/math?math=%5Cbegin%7Bequation%7D%20%5Clabel%7Beq%3Asur-err%7D%20E%5B%5Cepsilon_t%20%5Cepsilon'_%7Bt'%7D%5D%3D%5Csigma_%7Btt'%7D%20%5Cend%7Bequation%7D%20)

Note that this specification is very flexible, since it allows the coefficients ![$\beta_{it}$](https://render.githubusercontent.com/render/math?math=%24%5Cbeta_%7Bit%7D%24) to vary in order to modulate the effect of ![$X^i_{.t}$](https://render.githubusercontent.com/render/math?math=%24X%5Ei_%7B.t%7D%24) on ![$y_t$](https://render.githubusercontent.com/render/math?math=%24y_t%24). A less flexible if more parsimonious model can be obtained by imposing restrictions to some ![$\beta$](https://render.githubusercontent.com/render/math?math=%24%5Cbeta%24) coefficients to fix them across equations. This is equivalent to specifying an effect that is constant in time. In this case, we can reformulate the coefficients expression of ![$\beta_t = (\beta_{1}, \cdots, \beta_{r-1}, \beta_{r}, \beta_{r+1}, \cdots, \beta_{Nt})$](https://render.githubusercontent.com/render/math?math=%24%5Cbeta_t%20%3D%20(%5Cbeta_%7B1%7D%2C%20%5Ccdots%2C%20%5Cbeta_%7Br-1%7D%2C%20%5Cbeta_%7Br%7D%2C%20%5Cbeta_%7Br%2B1%7D%2C%20%5Ccdots%2C%20%5Cbeta_%7BNt%7D)%24) to restrict the first ![$r$](https://render.githubusercontent.com/render/math?math=%24r%24) coefficients to be constant.
-->

Equation \ref{eq:sur-sim}) can be rewriten in compact form as follows:

@@

@@

<!--
![\begin{equation} y = X \beta + \epsilon \end{equation}](https://render.githubusercontent.com/render/math?math=%5Cbegin%7Bequation%7D%20y%20%3D%20X%20%5Cbeta%20%2B%20%5Cepsilon%20%5Cend%7Bequation%7D)

\noindent where....
-->

As is the case with cross-sectional data, it is possible to test the residuals of Model \ref{eq:sur-sim} for spatial autocorrelation, and several tests have been developed to test the null hypothesis of spatial independence [see @Lopez2014spatial]. When the null hypothesis is rejected, several alternative specifications have been proposed to include spatial effects [@Anselin1988spatial; see also @Anselin2016estimation]. In this paper we consider a spatial SUR model that incorporates a spatial lag of the dependent variable as an explanatory factor. A sp

The stack expresion for the spatial lag SUR model (SUR-SLM),

@@

\noindent where $A =I_{TN}-\bf{\rho} \otimes W$ with $\bf{\rho} = diag(\rho_1, \cdots, \rho_T)$.

<!--
![\begin{equation} \label{eq:sur-slm} \bf{A}y = X \beta + \epsilon \\ \epsilon =N(0,\Omega) \end{equation}](https://render.githubusercontent.com/render/math?math=%5Cbegin%7Bequation%7D%20%5Clabel%7Beq%3Asur-slm%7D%20%5Cbf%7BA%7Dy%20%3D%20X%20%5Cbeta%20%2B%20%5Cepsilon%20%5C%5C%20%5Cepsilon%20%3DN(0%2C%5COmega)%20%5Cend%7Bequation%7D)

\noindent where ![$A =I_{TN}-\bf{\rho} \otimes W$ with $\bf{\rho} = diag(\rho_1, \cdots, \rho_T)$](https://render.githubusercontent.com/render/math?math=%24A%20%3DI_%7BTN%7D-%5Cbf%7B%5Crho%7D%20%5Cotimes%20W%24%20with%20%24%5Cbf%7B%5Crho%7D%20%3D%20diag(%5Crho_1%2C%20%5Ccdots%2C%20%5Crho_T)%24).
-->

This specification assumes that incidence in a province ($y_st$) at time $t$ is partially determined by the weighted average ($Wy_{st}$) of incidence in neighbouring provinces. Coefficients for the spatially lagged variable are estimated for each time period $\rho_t$ and identifies the intensity and the sign of the impacts of neighbourhood. It is possible test the null hypotheis of identtical levels of spatial dependence ($\rho_i=\rho_j, \forall i,j$). The correspond Wald test is available in the R package `spsur`.

<!--
This specification assumes that incidence in a province (![$y_st$](https://render.githubusercontent.com/render/math?math=%24y_st%24)) at time ![$t$](https://render.githubusercontent.com/render/math?math=%24t%24) is partially determined by the weighted average ![($Wy_{st}$)](https://render.githubusercontent.com/render/math?math=(%24Wy_%7Bst%7D%24)) of incidence in neighbouring provinces. Coefficients for the spatially lagged variable are estimated for each time period ![$\rho_t$](https://render.githubusercontent.com/render/math?math=%24%5Crho_t%24) and identifies the intensity and the sign of the impacts of neighbourhood. It is possible test the null hypotheis of identtical levels of spatial dependence (![$\rho_i=\rho_j, \forall i,j$](https://render.githubusercontent.com/render/math?math=%24%5Crho_i%3D%5Crho_j%2C%20%5Cforall%20i%2Cj%24)). The correspond Wald test is available in the R package `spsur`. 
-->

@@

Figure \ref{fig:weekly-average-incidence-map} shows the geographical variation in the incidence of the disease, as well as its temporal progression in weekly averages. It can be seen that the highest incidence at this early date was in the provice of Álava, in the North of Spain. While not the most populous province, with a population of only 331,549, Álava has the highest GDP per capita of all provinces. Vitoria, its main city, is the capital of the Basque country and has been the focus of efforts to develop a "Global Basque City" [@Meijers2008strategic], along with San Sebastian and Bilbao. The other early focus of the pandemic in Spain was Madrid, which is the most populous province in the country and has the second highest GDP per capita after Álava. The early outbreaks in these two provinces can be traced throughout the progression of the pandemic over time, although by the end of the period under consideration, other provinces had matched and even surpased their incidence rates, including Segovia and Soria to the north of Madrid, and Ciudad Real and Albacete to the south.

@@

```{r model1-8-day-moving-average, include=FALSE}
# Model with a lagged 8-day moving average of environmental variables:
sur.slm_lag8 <- spsur::spsurtime(formula = formula_lag8, 
                                 data=GPanel, 
                                 time = GPanel$Date, 
                                 type = "slm", 
                                 fit_method = "3sls", 
                                 listw=  listw,
                                 R = R2,
                                 b = b2)
#summary(sur.slm_lag8)
#print(paste("Pooled R^2 = ", sur.slm_lag8$R2[1]))
```

```{r model2-11-day-moving-average, include=FALSE}
# Model with 11-day moving average of environmental variables:
sur.slm_lag11 <- spsur::spsurtime(formula = formula_lag11, 
                                  data=GPanel, 
                                  time = GPanel$Date, 
                                  type = "slm", 
                                  fit_method = "3sls", 
                                  listw=  listw,
                                 R = R2,
                                 b = b2)
#summary(sur.slm_lag11)
#print(paste("Pooled R^2 = ", sur.slm_lag11$R2[1]))
```

```{r model3-11-day-weighted-moving-average, include=FALSE}
# Model with 11-day weighted moving average of environmental variables:
sur.slm_lag11w <- spsur::spsurtime(formula = formula_lag11w, 
                                   data=GPanel, 
                                   time = GPanel$Date, 
                                   type = "slm", 
                                   fit_method = "3sls", 
                                   listw=  listw,
                                 R = R2,
                                 b = b2)
#summary(sur.slm_lag11w)
#print(paste("Pooled R^2 = ", sur.slm_lag11w$R2[1]))
```
